\textls[-15]{\thetitle \mbox{(Technical Report)}}

### 2022-10-20 metatheory-condensed.tex
Save. Metatheory.

Now, we will mainly focus on the new typing rules that we introduced. For those curious about the complete formulation, we refer to our appendices or \citet{altkap2016, kaposi2017type} or maybe those not using QIIT \cite{coquand2018canonicity, sterling2019algebraic}. Our metatheory is in an extensional setting.

% linkage, signature and compilation and seal
$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}\quad
\Rule[name=Lnkg Type/Compile]
{\goodSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{\cL \sigma}{i}
\quad \goodType{\Gamma}{\cC \sigma}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\cCt o}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[\pi_1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
We start with signature (with a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is basically a list of types used to indicate the type of \textit{linkages}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript. Adding new types into signature \ruleref{Sig Add} requires a \textit{sealing judgement} $\goodSeal{\Gamma}{f}{\sigma}{A}$, which is an alias of a term judgement and will be explained later.  We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability.

@@

The sealing judgement provides a witness $f$ that a context $\Gamma, \cC\sigma$ can be abstracted into an \textit{interface} $\Gamma, A$ for a type $\goodType{\Gamma}{A}{}$. For example, sealing judgement can abstract an inductive type with constructors into an arbitrary type with functions (but without eliminators). Then any later fields $\goodTerm{\Gamma, A}{t}{T}$ relies only on the \textit{interface} $A$ can be inherited to other contexts (e.g. using extended inductive type), as long as the targeting contexts can be abstracted into the surrounding \textit{interface} for $t$ (i.e. $\Gamma, A$)  as well.

We now sketch out the importance of the compilation type. First we assume a singleton type (universe)\footnote{Something like this singleton type universe will appear later in inductive type formulation.} $A : S(T) \iff A \equiv T$, i.e. a type of one single type.  We can see that we have a term $x : \cC\{A : S(\bot), B : A \to \bot \} \vdash h_1 : \cC\{A : \cU, B:A \to \bot \}$ which is equivalent to $x : \Sigma (A: S(\bot)). A \to \bot \vdash h_1 : \Sigma (A: \cU). A \to \bot$. This can be considered as using seal to do an abstraction on $A:S(\bot)$.

But we can never have a term $x : \cL\{a : S(\bot), b : a \to \bot \} \vdash h_2 : \cL\{a : \cU, b : a \to \bot \}$, as we know $\cL\{a : S(\bot), b : a \to \bot \}$ is inhabited and that leads to a close term of $\cL\{a : \cU, b : a \to \bot \}$ due to the nature of late-binding of $\cL\sigma$---once we override $a$ with $\top$ we will have inconsistency using $b$.

@@

Our formulation of (extensible) inductive type is close to Wtype (\ruleref{WSig-Add} requires a pair of types $A \vdash B$ as well), but with mulitple constructors. Thus we have the signature of inductive type collecting each constructors, with a brand new judgement ${\goodWSig{\Gamma}{\tau}{n}}$. Unconventionally, a signature can induce a "universe" of types $\bW\tau$---in other words, $T : \bW\tau$ will be the inductive type instead of $\bW\tau$ itself. "Wsup" is used to construct a specific term of a given inductive type.

Similar to plugin implementation, we delegate handlers for recursion to linkages. We use "RecSig1" to compute the type of a single handler and "RecSig" for the complete family of handlers. Specifying the handler family using "Wrec" on target inductive term $w$ leads to elimination.

@@

Inheritance judgement can be further enriched with inheriting operation, overriding operation, extending operation and etc. These operations correspond to inherit, override and extend command inside "Family" during plugin implementation.

In this way, we will consider the program in our metatheory similar to those vanilla MLTT program, except there are meta variables denoting inheirtance judgement (i.e. "Family"), of which the only usage is to apply \ruleref{Inh}. We don't yet consider non-trivial equality between inheritance judgements.

@@

% Extending Inductive Type
We extend inductive type by overriding the inductive type with a enriched one. Thanks to sealing, most stuff from the parent family can be inherited to the children because enriched inductive type can export strictly stronger interface when abstracting.

@@

% Overriding + Pins
Finally, we pointed out how overridable and pins is modeled if we admit equality reflection. Despite the fact we have a complicated dynamic of invisibility between overridable and normal fields in the plugin, every field can be elaborated into the (pseudo)-syntax form \\ \mintinline{Coq}{Field hiding {x, y, z ...} newfield := t...}  where every former field x,y,z,...  hide its definition when type-check t, and remaining former fields have their definition exposed in t.
This above elaborated "hiding" syntax directly corresponding to the semantic of sealing---when definition is exposed, we expose propositional equality into the context, which will be a judgemental equality with the help of equality reflection; otherwise we only expose the type and no extra information.

>>>>

Now, we will mainly focus on the newly introduced typing rules, and illustrate their connection to the plugin implementation. For those curious about the complete meta-theory formulation, we refer to our appendices or \citet{altkap2016, kaposi2017type} or maybe those not using QIIT \cite{coquand2018canonicity, sterling2019algebraic}. Our metatheory is in an extensional setting.

% linkage, signature and compilation and seal
$$
\Rule[name=Ept Sig]
{}
{\goodSig{\Gamma}{\nu\cdot}{0}}\quad
\Rule[name=Lnkg Type/Compile]
{\goodSig{\Gamma}{\sigma}{n}}
{\goodType{\Gamma}{\cL \sigma}{i}
\quad \goodType{\Gamma}{\cC \sigma}{}}
\quad
\Rule[name=Lnkg Compile]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
}
{
  \goodTerm{\Gamma}{\cCt o}{\cC \sigma}
}
$$
$$
\Rule[name=Sig Add]
{\goodSig{\Gamma}{\sigma}{n} 
 \quad \goodType{\Gamma}{A}{}
 \quad \goodSeal{\Gamma}{f}{\sigma}{A}
 \quad \goodType{\Gamma, A}{T}{i}}
{\goodSig{\Gamma}{(\nu^+ \ \sigma \ \{f\} \ T)}{n+1}}
\quad
\Biggl( \boxed{\goodSeal{\Gamma}{s}{\sigma}{A}} 
:= \boxed{\goodTerm{\Gamma, \cC\sigma}{s}{A[\pi_1]}} \Biggr)
$$
$$
\Rule[name=Ept Lnkg]
{}
{\goodTerm{\Gamma}{\mu\cdot}{\cL \nu\cdot}}
\quad
\Rule[name=Lnkg Add]
{ \goodTerm{\Gamma}{o}{\cL \sigma} 
\quad \goodType{\Gamma}{A}{}
\quad  {\goodSeal{\Gamma}{f}{\sigma}{A}}
 \quad \boxed{\goodTerm{\Gamma, A}{t}{T}}
}
{\goodTerm{\Gamma}{(\mu^+ \ o \ \{f\} \ t)}{\cL(\nu^+ \ \sigma \ \{f\} \ T)}}
$$
We start with signature (in a brand new judgement $\goodSig{\Gamma}{\sigma}{n}$). A signature is a list of types contained by a \textit{linkage}\footnote{Following the convention by \citet{zm2017}, we name the meta-theoretical correspondents of \textit{family} as \textit{linkage}.}, tracking the length as superscript.   We will use $\{a : T_1, b : T_2, ..\}$ as a shorthand for signature with type $T_1, T_2, ..$ for readability.

Adding new types into signature \ruleref{Sig Add} requires a \textit{sealing judgement} $\goodSeal{\Gamma}{f}{\sigma}{A}$, which is an alias of a term judgement. Intuitively, sealing can abstract any given type by indicating a morphism to the interface. Sealing is a convenient device when programming using meta-theory, \textit{to indicate whether each field is overridable, non-overridable (thus definitionally exposed) or extensible}. For example, if a field is overridable, we abstract that to its type; if a field has exposed its definition, we abstract it into the singleton type containing it instead.

However, there is subtle difference compared to the plugin implementation---$(\mu^+ ~o~\{f\}~t)$ reads as \textit{``$t$ is a new well-typed field when we abstract the \textbf{prior} fields using $f$''}. In other words, every field in meta-theory has a right to choose how they perceive and abstract the prior fields. While in the plugin, "Overridable" will decide how the current defining field is percived by the following fields.

@@

% The sealing judgement provides a witness $f$ that a context $\Gamma, \cC\sigma$ can be abstracted into an \textit{interface} $\Gamma, A$ for a type $\goodType{\Gamma}{A}{}$. For example, sealing judgement can abstract an inductive type with constructors into an arbitrary type with functions (but without eliminators). Then any later fields $\goodTerm{\Gamma, A}{t}{T}$ relies only on the \textit{interface} $A$ can be inherited to other contexts (e.g. using extended inductive type), as long as the targeting contexts can be abstracted into the surrounding \textit{interface} for $t$ (i.e. $\Gamma, A$)  as well.

The importance of the compilation type lies in its difference from the linkage type. First we have a singleton type $"s A" : S(T) \iff "A" \equiv T$, i.e. a type of one single term.  We can see that we have a term $x : \cC\{A : S(\bot), B : A \to \bot \} \vdash h_1 : \cC\{A : \cU, B:A \to \bot \}$ which is equivalent to $x : \Sigma (A: S(\bot)). A \to \bot \vdash h_1 : \Sigma (A: \cU). A \to \bot$. This can be considered as using seal to do an abstraction on $A:S(\bot)$.

But we can never have a term $x : \cL\{a : S(\bot), b : a \to \bot \} \vdash h_2 : \cL\{a : \cU, b : a \to \bot \}$, as we know $\cL\{a : S(\bot), b : a \to \bot \}$ is inhabited and that leads to a close term of $\cL\{a : \cU, b : a \to \bot \}$ due to the nature of late-binding of $\cL\sigma$---once we override $a$ with $\top$ we will have inconsistency using $b$. This example illustrates the programming difference between the sigma type and linkage type.

@@

Our formulation of (extensible) inductive type is close to W-type (\ruleref{WSig-Add} requires a pair of types $A \vdash B$ as well), but with mulitple constructors. Thus we have a brand new judgement ${\goodWSig{\Gamma}{\tau}{n}}$ indicating signature of inductive type by its constructors. Unconventionally, a signature can induce a "universe" of types $\bW\tau$---in other words, $T : \bW\tau$ will be the inductive type instead of $\bW\tau$ itself. "Wsup" is used to construct a specific term of a given inductive type.

When programming via meta-theory, we usually seal the inductive type, only expose its constructors (as a mundane function member) and hide the eliminator by abstracting $T : \bW\tau$ into $"W"~T : \cU$. Then the later fields---those only rely on this interface with $\cU$---can be inherited to other contexts (e.g. using enriched inductive type), as long as the targeting contexts can be abstracted into this {interface} again.

Similar to plugin implementation, we delegate handlers (and their inheritance structure) for recursion to linkages. We use "RecSig1" to compute the type of a single handler and "RecSig" for the complete family of handlers for all constructors. \ruleref{Rec-Constr} uses "RecSig" to achieve exhaustiveness checking on elimination. The partial recursors and the computational axioms are treated as normal fields as expected, since we know they are provable in the plugin implementation.

@@

Inheritance judgement can be further enriched with inheriting operation, overriding operation, extending operation and etc. These operations correspond to inherit, override and extend command inside "Family" in our plugin implementation.

In this way, we will consider the program in our metatheory similar to those vanilla MLTT program, except there are meta variables denoting inheirtance judgement (i.e. "Family"), of which the only usage is to apply \ruleref{Inh}. We don't yet consider non-trivial equality between derivations of inheritance judgements.

@@

% Extending Inductive Type

% Focus on FRecursion and FInduction

@@

% Overriding + Pins
Finally, we pointed out how overridable and pins is modeled.. Despite the fact we have a complicated dynamic of invisibility between overridable and normal fields in the plugin, every field can be elaborated into the (pseudo)-syntax form \\ \mintinline{Coq}{Field hiding {x, y, z ...} newfield := t...}  where every former field x,y,z,...  hide its definition when type-check t, and remaining former fields have their definition exposed in t.
This above elaborated "hiding" syntax directly corresponding to the semantic of sealing---when definition is exposed, we expose a term by using singleton type, which will provide us a judgemental equality to the definition.

### 2022-10-20 syntax.tex
Save. Metatheory.

Finally, we make a note here on how sealing can easily express the overidability in our implementation. To expose the definition in our metatheory, we model that by adding equality reflection into our metatheory (and thus change our setting into extensional type theory). Then we can consider exposing the concrete definition as the existence of an identity type in the context. The key insight of the sealing in our metatheory, is that it gives the user the ability to manually specify which former fields expose its definition and which  former fields only expose their types---for those only expose their types, the corresponding identity type will be relaxed into a unit type. Now look back at the overridability mechanism in our implementation, it is a much restricted version: each overridable field can only see the concrete definition of the former non-overridable(default) fields and depending overridable fields; for each non-overridable(default) field, only the concrete definition of former non-overrdiable(default) fields are exposed.

>>>>

Finally, we make a note here on how sealing can easily express the overidability in our implementation. To expose the definition in our metatheory, we model that by using singleton type, where $"s A" : S(T) \iff "A" \equiv T$ . The key insight of the sealing in our metatheory, is that it gives the user the ability to manually specify which former fields expose its definition and which  former fields only expose their types---for those only expose their types, the corresponding identity type will be relaxed into a unit type. Now look back at the overridability mechanism in our implementation, it is a much restricted version: each overridable field can only see the concrete definition of the former non-overridable(default) fields and depending overridable fields; for each non-overridable(default) field, only the concrete definition of former non-overrdiable(default) fields are exposed.
